---
title: "Increased aridity is associated with stronger tradeoffs in ponderosa pine vital functions"
author: "Angie Gonzalez"
date: '2023-01-28'
output: html_document
---

This formatted markdown document includes all analyses described in the following article:
Gonzalez, A.D., Pearse, I.S., Redmond, M.D. Increased aridity is associated with stronger tradeoffs in ponderosa pine vital functions.


The code below is separated into three groupings: 1) model selection, 2) models over time, and 3) models among individuals. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading necessary packages for analysis
```{r packages, echo=F, message=F, warning=F}

install.packages(c("sjPlot", "tidyverse", "lme4", "glmmTMB", "lmerTest", "ggeffects", "MuMIn", "effects"))

Packages <- c ("tidyverse", "lme4", "sjPlot", "glmmTMB", "lmerTest", "ggeffects", "effects", "MuMIn")
lapply(Packages, library, character.only = TRUE)

```

## Loading files and formatting
```{r files, echo=T}

all_data <- read.csv("annual_data.csv")
all_data <- all_data %>% subset(select=-c(X)) #dropping unneccessary columns
tree_avgs <- read.csv("averages_data.csv")

ggplot(all_data)+geom_bar(aes(year)) # reviewing cone counts by year
all_data <- all_data %>% subset(year<=2020 & year>=2003) # dropping 2003 and prior due to limited observations

```
# Model Selection ########
```{r model-fitting, echo=F}

## which resin metric makes most sense to use? AIC test to determine --- 

test <- lmer(resinarea2~growth12+(1|site/tree),na.action = "na.fail", all_data) # the best by far
test1 <- lmer(totalresinarea~growth12+(1|site/tree),na.action = "na.fail", all_data)
test2 <- lmer(meanductsize~growth12+(1|site/tree),na.action = "na.fail", all_data)
AICc(test, test1, test2)

# one more time w/ cones
test <- lmer(resinarea2~conematur+(1|site/tree),na.action = "na.fail", all_data) # again - the best by far
test1 <- lmer(totalresinarea~conematur+(1|site/tree),na.action = "na.fail", all_data)
test2 <- lmer(meanductsize~conematur+(1|site/tree),na.action = "na.fail", all_data)
AICc(test, test1, test2)

# now lets move to testing the appropriate predictors and controls

test <- lmer(resinarea2~growth12+FDSI+(1|site/tree),na.action = "na.fail", all_data)
test1 <- lmer(resinarea2~growth12+BA5+dbh+(1|site),na.action = "na.fail", tree_avgs)

dredge(test) # best model w/out FDSI
dredge(test1) # best model w/out BA or DBH

#checking again w/ other combos just to be safe. Will use same test df names to overwrite.
test <- lmer(growth12~conematur+FDSI+(1|site/tree), na.action = "na.fail", all_data)
test1 <- lmer(growth12~conematur+BA5+dbh+(1|site), na.action = "na.fail", tree_avgs)


dredge(test) # best model w/ FDSI
dredge(test1) # best model shows nothing. Pretty much saying its a bad model. 

test <- lmer(resinarea2~conematur+FDSI+(1|site/tree), na.action = "na.fail", all_data)
test1 <- lmer(resinarea2~conematur+BA5+dbh+(1|site), na.action = "na.fail", tree_avgs)

dredge(test) # cone mature and FDSI are second to just FDSI, aic difference is 2 so i feel ok sticking with both.
dredge(test1) 
# model with only FDSI. Our interest is in tradeoffs among vital functions so we will be keeping conematur regardless.


remove(test, test1)

````

```{r exploration-colinearity}


# NOTE: the outlier below was removed in the data calculations RMD script before the tree averages csv was created. I found a few outliers in the averages data frame but did not remove them because of the small number of observations, the fact that certain trees were not consistent outliers across variables makes it difficult to decide to remove outlier. The only outlier that was grossly higher than the others is the one below. I may also be doing transformations on these models to meet assumptions; Following assumption checks, outliers, colinearity review as suggested in Zurr 2009.

# outlier was found in resinarea column. double checking - resin area ~11.
all_data[785,] # our outlier is in row 785
all_data <- all_data[-785,] # pretty huge outlier. choosing to remove data.

#checking correlations#
datapair <- all_data
datapair <- datapair %>% na.omit()
datapair <- datapair %>% subset(select=-c(year, tree, site, growth5, slope, aspect, dbh, crown.class, height, crown.fullness, BA10, BA20, Leadershoots, wue, cones))
datapair <- lapply(datapair, as.numeric) # ensuring all are reading as numeric
pairs(datapair)
# looks good. 
datapair <- as.data.frame(datapair)
datacor<-as.data.frame(cor(datapair))
datacor
# all looks fine. 
remove(datacor, datapair)
```


### Models Among Individuals ############################################

The models below are all analyzed using the "tree_avgs" CSV.

Final models will be as follows y~x* (cwd or wue) + (1|site)
results from AIcc. 

resin~growth models among individuals
```{r resin-growth, echo=F}

dotchart(log(tree_avgs$resinarea2))
dotchart(tree_avgs$resinarea2) # log looks much better

## C13 and CWD models for resin~growth relationship among individuals (avgs)
tree_ductgrow_mod1 <-lmer(log(resinarea2)~scale(growth12)*scale(wuecentered)+(1|site), tree_avgs)
tree_ductgrow_mod2 <-lmer(log(resinarea2)~scale(growth12)*scale(cwd)+(1|site), tree_avgs)

plot_model(tree_ductgrow_mod1, type="diag") #not the best, but not terrible. looks OK
plot_model(tree_ductgrow_mod2, type="diag") #OK

summary(tree_ductgrow_mod1) # growth is positively correlated. No other term is significant. Carbon discrimination -no impact.
tab_model(tree_ductgrow_mod1, show.df = TRUE, show.stat = TRUE) ## this output gives us an html table with our standardized estimates, CIs and other summary outputs

summary(tree_ductgrow_mod2) # same pattern as above. Only growth as positive correlation. Nothing else significant. 
tab_model(tree_ductgrow_mod2, show.df=TRUE, show.stat=TRUE)

```

Plotting significant models 

```{r resin-growth-plots, echo=F, include=F}
# Let's plot the first model w/ out the interaction since it was not significant. If interaction is not significant we will plot the first model ran (c13) for consistency. 

# making data frame with resin-growth estimates pulled from model
ef_growth <-ggpredict(tree_ductgrow_mod1, terms = c("growth12"), type = "fixed")
ef_growth

#making plot ~ 

ggplot(ef_growth, aes(x = x, y = predicted)) +
  geom_line(linetype="F1")+
  geom_ribbon( aes(ymin = conf.low, ymax = conf.high), alpha = .15) + geom_point(data=tree_avgs, aes(growth12, resinarea2),
  ) +labs(title="", shape="Site", y="Average Relative Duct Area (% annual ring)", x="Average Annual Growth (mm)"
  ) + theme_gray(base_size=9) 

```

resin~cone models among individuals

```{r resin-cones}

tree_ductscone <- lmer(log(resinarea2)~scale(conematur)*scale(wuecentered)+(1|site), tree_avgs) 
plot_model(tree_ductscone, type="diag") # not bad. Some issues, but nothing major
dotchart(log(tree_avgs$resinarea2)) # definitely should be log transformed. any potential outliers have far less influence because spread is increased substantially.  

tree_ductscone1 <- lmer(resinarea2~scale(conematur)*scale(wuecentered)+(1|site), tree_avgs) 
plot_model(tree_ductscone1, type="diag") ## the back transformation makes the it slightly worse, it has a slight skew without it. let's keep it transformed so it has the bell curve and the spread is a little better for homogeneity. 
remove(tree_ductscone1)

# cwd
tree_ductscone1 <- lmer(log(resinarea2)~scale(conematur)*scale(cwd)+(1|site), tree_avgs) 

plot_model(tree_ductscone, type="pred")
# summary output of both models
summary(tree_ductscone)
tab_model(tree_ductscone, show.stat=TRUE)
summary(tree_ductscone1)
tab_model(tree_ductscone1, show.stat = TRUE)


````

plotting significant models. Plotting only resin~carbon model

```{r resin-carbon-plotting}

ef_ductcone13 <-ggpredict(tree_ductscone, terms = c("wuecentered"), type = "fixed")

ggplot(ef_ductcone13, aes(x = x, y = predicted, stroke=0))+ geom_point(data=tree_avgs, aes(x=wuecentered, y=resinarea2, stroke=0),color="gray",
)+geom_line(size=1)+geom_ribbon( aes(ymin = conf.low, ymax = conf.high), alpha = .15)  +labs(title="", shape="Site", y="Average Relative Duct Area (% annual ring)", x=expression(Delta*"13C - Centered by Site")
  ) + scale_shape_manual(values=1:8)+theme_bw(base_size=9)+annotate(geom="text", y=0.9, x=-1.4, label="(a)",size=3,family="sans"
)


## because our centered carbon discrimination values are between -1 and 1, i limited the x axis so it looks better visually. 

ef_ductcone <-ggpredict(tree_ductscone, terms = c("conematur"), type = "fixed")
#making plot ~ 
ggplot(ef_ductcone, aes(x = x, y = predicted), stroke=0) +geom_point(data=tree_avgs, aes(x=conematur, y=resinarea2),color="gray",stroke=0)+ geom_line()+geom_ribbon( aes(ymin = conf.low, ymax = conf.high), alpha = .15) +labs(title="", shape="Site", y="", x="Average Cone Abundance") + scale_shape_manual(values=1:8)+theme_bw(base_size=9)+annotate(geom="text", y=0.9, x=0, label="(b)",size=3,family="sans")

```

moving forward to the final model among individuals model.
```{r growth-cones}
# we are going to try this log transformed since cones is in our y axis.


tree_conesgrowth <- lmer(log(growth12)~scale(conematur)*scale(wuecentered)+(1|site), tree_avgs)

plot_model(tree_conesgrowth, type="diag") # looks good.

# trying w/out log transformation
tree_conesgrowth1 <-lmer(growth12~scale(conematur)*scale(wuecentered)+(1|site), tree_avgs)
plot_model(tree_conesgrowth1, type="diag") # worse- but nothing too crazy. let's remove this and stick with log transformation
# double checking spread
dotchart(log(tree_avgs$growth12))
dotchart(tree_avgs$growth12) # keeping it transformed. 

remove(tree_conesgrowth1)

#CWD interaction:
tree_conesgrowth1 <-lmer(log(growth12)~scale(conematur)*scale(cwd)+(1|site), tree_avgs)


summary(tree_conesgrowth)
tab_model(tree_conesgrowth, show.stat = TRUE)
summary(tree_conesgrowth1)
tab_model(tree_conesgrowth1, show.stat=TRUE)


ef_conegrow <-ggpredict(tree_conesgrowth, terms = c("conematur"), type = "fixed")

#conegrow <-
ggplot(ef_conegrow, aes(x = x, y = predicted, stroke=0,size=0.00005)) + geom_point(data=tree_avgs, aes(x=conematur, y=resinarea2), color="gray",
) + geom_line(size=1)+geom_ribbon( aes(ymin = conf.low, ymax = conf.high), alpha = .15) +labs(title="", y="Average Relative Duct Area (% annual ring)", x="Average Cone Abundance"
  ) + theme_bw(base_size=9)+ theme(legend.position = "none")


```
Nothing to be plotted in this last set of analyses. We aren't seeing any trade offs among functions among individuals. We did see a positive correlation between resin and growth. In that individuals who had higher growth also had higher resin. Carbon discrimination did impact average resin area. Individuals with higher carbon discrimination had lower average resin area. 

### Models over time###################################

Let's move forward with our analyses over time. This is probably where we will start seeing some interesting things. These models are divided up in the same fashion. The main question with these analyses are: How is an individual varying its resource allocation patterns annually? Are we seeing different trade off patterns over time and is climate impacting this?  

```{r resin-growth}
#WUE model:
year_ductgrow_mod <-lmer(resinarea2~growth12*wuecentered+(1|site/tree), all_data)

plot_model(year_ductgrow_mod, type="diag") # not great, but not bad. Let's try sqrt transformation. Heterogeneity is a concern here.

year_ductgrow_mod1 <-lmer(sqrt(resinarea2)~(growth12)*scale(wuecentered)+(1|site/tree), all_data) # sqrt transformation because of 0 values.
plot_model(year_ductgrow_mod1, type="diag") # significantly better; the spread is increased. Heterogeneity is still of concern but it doesnt seem to be a major violation. Most of it has a good spread. Some clustering at the left but nothing too crazy. linearity and normality look pretty good. some stuff at tail end. 
dotchart(all_data$resinarea2)
dotchart(sqrt(all_data$resinarea2)) # better with transformation; there is a wider spread but you can still see several potential outliers. hard to tell here. 

remove(year_ductgrow_mod)

summary(year_ductgrow_mod1)

tab_model(year_ductgrow_mod1, show.df=TRUE, show.stat = TRUE) 

#CWD model:
year_ductgrow_mod2 <- lmer(sqrt(resinarea2)~scale(growth12)*scale(cwd)+(1|site/tree), all_data)

summary(year_ductgrow_mod2)
tab_model(year_ductgrow_mod2, show.stat = TRUE, show.df = TRUE)

```

plotting model
```{r resin-growth-plotting}

### you are going to need to redo this - i will make a note to myself so i remember to fix it. 

ef_growth1 <-ggpredict(year_ductgrow_mod1, terms = c("growth12"), type = "fixed")
ef_growth1


  ggplot(ef_growth1, aes(x = x, y = predicted, stroke=0, size=0.00005))+ geom_jitter(data=all_data, aes(x=growth12, y=resinarea2, stroke=0, size=0.0000001), color="gray",
  ) + geom_line(size=1)+geom_ribbon( aes(ymin = conf.low, ymax = conf.high), alpha = .15) +labs(title="", y="Relative Duct Area (% annual ring)", x="Annual Growth (mm)"
) + theme_bw(base_size=9) +  theme(legend.position = "none") +annotate(geom="text", y=1.9, x=0, label="(a)",size=3,family="sans"
)


ef_cwd_Defgrow <-ggpredict(year_ductgrow_mod2, terms = c("growth12", "cwd[meansd]"))

 plot(ef_cwd_Defgrow, add.data=TRUE,colors=c("#8fb771","#8961b3", "#b8545c"), dot.alpha=.5, dot.size=0.5
            ) + labs(title = "", x="Annual Growth (mm)", y="Relative Duct Area (% annual ring)", color="CWD (mm)"
            ) + theme(legend.position = "top", text=element_text(size=9)) 

```


moving on to the resin-cone relationship. 
```{r resin-cone}

#wue model#

#let's double check the transformation on this one as well
year_ductscone <- lmer(resinarea2~scale(conematur)*scale(wuecentered)+scale(FDSI)+(1|site/tree), all_data)
plot_model(year_ductscone, type="diag") # very bad heterogenity violation

year_ductscone1 <- lmer(sqrt(resinarea2)~scale(conematur)*scale(wuecentered)+scale(FDSI)+(1|site/tree), all_data) 
plot_model(year_ductscone1, type="diag") 
## once again, not much as changed. Diag is not bad, but definitely not perfect. Will go w/ sqrt transformation again because it slightly increased that spread of the normal distribution.
remove(year_ductscone)

summary(year_ductscone1)
tionship. no interaction, FDSI is behaving expected.

#cwd model#
year_ductscone <- lmer(sqrt(resinarea2)~scale(conematur)*scale(cwd)+scale(FDSI)+(1|site/tree), all_data) 
plot_model(year_ductscone, type="diag") 

summary(year_ductscone)

tab_model(year_ductscone, show.stat=TRUE, show.df = TRUE)
tab_model(year_ductscone1, show.stat=TRUE, show.df = TRUE)


````

```{r resin-cone-plotting}


ef_cone <-ggpredict(year_ductscone1, terms = c("conematur"), type = "fixed")
# let's plot these in the same manner as we have been for consistency.


  ggplot(ef_cone, aes(x = x, y = predicted, stroke=0,size=0.00005)) + geom_jitter(data=all_data, aes(x=conematur, y=resinarea2, stroke=0, size=0.0000001), color="gray",
) + geom_line(size=1)+geom_ribbon( aes(ymin = conf.low, ymax = conf.high), alpha = .15) +labs(title="", y="Relative Duct Area (% annual ring)", x="Annual Cone Abundance"
  ) + theme_bw(base_size=9)+ theme(legend.position = "none")+annotate(geom="text", y=1.9, x=1, label="(b)",size=3,family="sans"
  )


## the resin wue plot
ef_ductwue <- as.data.frame(effect(term =c("scale(wuecentered)"), mod=year_ductscone1))

summary(ef_ductwue) # double checking it looks correct. 

ggplot() + geom_point(data=all_data, aes(x= wuecentered, y=sqrt(resinarea2)), color="darkgrey"
) +geom_line(data=ef_ductwue, aes(x=wuecentered, y=fit), color="black", linetype="F1"
) +geom_ribbon(data= ef_ductwue, aes(x=wuecentered, ymin=lower, ymax=upper),alpha= 0.3
) +labs(shape="Site", y=expression(sqrt("Relative Duct Area ") *" (% annual ring)"), x=expression(Delta*"13C - Centered by Site")
) + xlim(-1, 1)+ theme_gray(base_size=9) +theme_bw()
````

Moving on to our final model. Growth and cones.
```{r growth-cone}
# let's double check our log transformation. Based on what we saw above I am assuming we will need it to be log transformed. 
# we are going to play with some transformations on this one. I am going to try sqrt first. 
yr_conegrow<- lmer(growth12~scale(conematur)*scale(wuecentered)+scale(FDSI)+(1|site/tree), data=all_data)

plot_model(yr_conegrow, type="diag") # once again, heterogeneity violated. Sqrt?
yr_conegrow1<- lmer(sqrt(growth12)~scale(conematur)*scale(wuecentered)+scale(FDSI)+(1|site/tree), data=all_data)
plot_model(yr_conegrow1, type="diag") # better but not great. there is still a small fan pattern.
yr_conegrow2<- lmer(log(growth12+1)~scale(conematur)*scale(wuecentered)+scale(FDSI)+(1|site/tree), data=all_data) # +1 for 0s
plot_model(yr_conegrow2, type="diag") # it looks the same. i'll stick with sqrt for consistency. 

remove(yr_conegrow, yr_conegrow2)

dotchart(all_data$growth12) # one more check of spread 
dotchart(sqrt(all_data$growth12)) # def needs transformation

summary(yr_conegrow1)

yr_conegrow2<- lmer(sqrt(growth12)~scale(conematur)*scale(cwd)+scale(FDSI)+(1|site/tree), data=all_data)
plot_model(yr_conegrow2, type="diag") # just a double check
# this one is tough. tranformed it does improve normality and linearity, it also increases spread for homoscedacity but there is still evidence of violations of heterogenity. others look fine though. Let's keep it. 

summary(yr_conegrow2) 


tab_model(yr_conegrow1, show.stat=TRUE, show.df=TRUE)
tab_model(yr_conegrow2, show.stat=TRUE, show.df=TRUE)


```

```{r growth-cones-plotting}

# 
plot_model(yr_conegrow2, type="int", mdrt.values = "meansd", title=c(""
    ),legend.title =c("Climatic Water Deficit"),axis.title=c("Cone Abundance", "Annual Growth (mm)"
    ), group.terms = c(1, 2, 3), colors = "bw", show.data=TRUE, dot.size=0.5
  ) +theme_bw() # nice. Interesting relationship here. The lowest CWD sites seem to have a very small trade off. It is almost parallel. 

ef_cwdconegrow <- ggpredict(yr_conegrow2, terms = c("conematur", "cwd[meansd]"))
ef_cwdconegrow

  plot(ef_cwdconegrow, add.data=TRUE, colors=c("#8fb771","#8961b3", "#b8545c"), dot.alpha=.5, dot.size=0.5
       ) + labs(title="", x="Annual Cone Abundance", y="Annual Growth (mm)"
    ) + theme(legend.position = "none",  text=element_text(size=9)) +annotate(geom="text", x=15, y=6, label="(b)",size=3,family="sans",
    )



```


